% Naija Persona Agent — Hackathon Solution Paper
% Compile: pdflatex paper.tex && bibtex paper && pdflatex paper.tex && pdflatex paper.tex

\documentclass[sigconf,nonacm,balance]{acmart}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{booktabs}

\title{The Cultural Prior in LLM Agents:\\
       An Open-Source Recovery for Nigerian Review Simulation and Recommendation}

\author{Ashinze Ifeanyi}
\affiliation{\institution{Naija Persona Agent Team}\country{Nigeria}}
\email{ashinze@bluebulb.co.uk}

\author{Franca}
\affiliation{\institution{Naija Persona Agent Team}\country{Nigeria}}

\author{Anonymous Third Author}
\affiliation{\institution{Naija Persona Agent Team}\country{Nigeria}}

\begin{document}

\begin{abstract}
Large language model (LLM) agents trained on predominantly Western web corpora
carry an implicit \textbf{cultural prior}: when prompted to simulate a Nigerian
user, vanilla frontier models compress rating intensity toward the centre,
flatten Pidgin and Nigerian-English register into a sanitized international
voice, default to individualistic framing, and routinely misread religious
markers (``By God's grace'', ``Alhamdulillah'') as customer-service complaints.
We present \textbf{Naija Persona Agent} (NPA), a two-task agent system that
makes the cultural prior visible and recovers it through three coordinated
mechanisms: (1) a structured cognitive persona representation with a
Nigerian register tier, (2) register-aware retrieval-augmented prompting,
and (3) \textbf{NaijaReviewer-8B}, an open-weight Llama 3.1 8B QLoRA
fine-tune trained on \(\sim\!20\)k Nigerian product reviews that we release
with the paper. On a fresh 100-row held-out test split (generated by Llama-3.3-70B, a
different model than what generated our training corpus, eliminating
single-generator confounds), NaijaReviewer-8B beats Claude Sonnet 4 on
the metric that matters most for the rubric's rating-accuracy line:
\textbf{RMSE 1.114 vs 1.319 ($-$15.5\,\%)}. The gap is confirmed by
the external AgentSociety arbiter (\texttt{preference\_estimation}
0.792 vs 0.784) and holds on 78\,\% out-of-training personas. On
text-quality metrics the two backbones tie (BERTScore F1 0.858 vs
0.857; AgentSociety \texttt{overall\_quality} 0.787 vs 0.795, within
0.008). Our cultural-marker and register-tier metrics show large
nominal wins (53\,\% vs 35\,\% register match, 13\,\% vs 7\,\% marker
recall), but these are partially confounded --- 69\,\% of v2 GT markers
overlap with the explicit guidance in our prompt template, so the
metric is reading as marker \emph{emission rate}, not authenticity
(§\ref{sec:audit}). Task B (recommendation re-ranking) is a clean
sweep on a smaller sample: \textsf{NDCG@10} \textbf{0.623 vs 0.485
($+$28\,\%)}, \textsf{HR@5} 0.647 vs 0.431. The fine-tune attains all
of this while being 17$\times$ smaller (8\,B vs 70\,B baseline) and
running locally on a single consumer Mac at zero marginal API cost. The system, model weights,
training corpus, and Streamlit demo are released under permissive
licenses.
\end{abstract}

\maketitle

% ---------------------------------------------------------------
\section{Introduction}

When a Lagos-based small-business owner writes ``e too much abeg, but value for
money no go lie'' on a marketplace review, three things are happening at once.
First, the writer is code-switching between Nigerian Pidgin and English at
sub-clause granularity. Second, the writer is using a culturally specific
intensity calibration --- ``too much'' here means \emph{a lot of value}, not
\emph{excessive}. Third, the writer is framing the purchase decision against an
implicit communal reference (``value for money'' = is this respectable to bring
home), not against an individual one. Vanilla frontier LLMs, when prompted to
simulate this user, fail at all three: they straighten the code-switching into
formal English, mistranslate ``too much'' as a complaint, and default to an
individualistic value framing.

We call this gap the \textbf{cultural prior} of the model. The prior is real,
quantifiable, and --- crucially --- recoverable. This paper is a 5-day,
3-person sprint to recover it for the Nigerian context, in service of two
tasks defined by the AgentSociety / Nigerian AI Agents Hackathon:
\textbf{Task~A} (User Modeling: given a persona and a product, simulate the
review and rating the user would write); and \textbf{Task~B} (Recommendation:
given a persona, rank products from a candidate set).

\textbf{Contributions.}
\begin{itemize}[leftmargin=*]
    \item A \emph{structured cognitive persona representation} that decomposes
          a user into four dimensions (hedonic/utilitarian, communal/individual,
          intensity calibration, aspect priorities) plus a four-level Nigerian
          register tier --- a representation portable across both tasks and
          across markets (the same schema extends to Ghana, Kenya, South Africa).
    \item \emph{NaijaReviewer-8B}: a QLoRA fine-tune of Llama 3.1 8B Instruct
          on \(\sim\!20\)k Nigerian product reviews, released under the Llama
          3.1 Community License. \(Q4\_K\_M\) GGUF and merged FP16 weights are
          on HuggingFace. Runs on a single consumer Mac via LM Studio.
    \item Explicit \emph{cold-start}, \emph{cross-domain}, and \emph{multi-turn}
          recommendation pathways --- the three Task-B sub-scenarios named in
          the brief. The agent surfaces its own scenario detection
          (\textsf{cold\_start} / \textsf{cross\_domain} / \textsf{multi\_turn}
          flags) and a narrative reasoning trace explaining each decision.
    \item A \emph{paper-first evaluation suite} covering both tasks against
          three baselines (vanilla Claude Sonnet 4, vanilla GPT-4o, base Llama
          3.1 8B), with bootstrap-CI'd RMSE / BERTScore F1 / ROUGE-L /
          register-match / cultural-marker recall, AND the official
          AgentSociety simulator metrics (\textsf{preference\_estimation},
          \textsf{sentiment\_error}, \textsf{topic\_error},
          \textsf{emotion\_error}, \textsf{overall\_quality}).
    \item An \emph{open data release}: 24 hand-curated Nigerian personas
          spanning 6 geopolitical zones $\times$ 4 age bands $\times$ 4 register
          tiers $\times$ 3 religions, plus a filtered 6\,657-product real-Jumia
          catalogue. All released CC-BY-4.0.
    \item A \emph{product roadmap} (§\ref{sec:product}): NaijaPersona ---
          eight commercial SKUs built on top of the same engine, targeted at
          African brand / marketing / retention teams currently overpaying for
          Western customer-research tooling that misreads their users.
    \item A \emph{reproducible open-source system}: FastAPI service, Streamlit
          demo, fine-tuning notebook, eval harness --- everything one command
          from a fresh clone.
\end{itemize}

% ---------------------------------------------------------------
\section{Related Work}

\subsubsection*{LLM user simulation.} AgentCF++~\cite{agentcf} and
RecAgent~\cite{recagent} both model users with concatenated rating histories
and natural-language profiles. Our cognitive decomposition is closer in
spirit to USHB and the Tsinghua AKF entry from the 2024 AgentSociety
Challenge, which used structured behavioural priors to drive user simulation;
we extend this with an explicit cultural register dimension.

\subsubsection*{Nigerian NLP.}  AfriSenti~\cite{afrisenti} and
NaijaSenti~\cite{naijasenti} ship sentiment-labelled corpora for Hausa,
Igbo, Yoruba, and Pidgin tweets; SentiLeye~\cite{sentileye} provides Yoruba
movie review sentiment.  Lin~\cite{lin2024} shows that fine-tuning small
LLMs on Nigerian dialogue recovers \(>\!60\)\% of frontier-model quality at
\(<\!5\)\% of the inference cost. AfriBERTa~\cite{afriberta} and the
Masakhane community~\cite{masakhane} have produced the canonical
African-language encoder/decoder pretraining checkpoints. We build on all of
these as data sources for our training corpus.

\subsubsection*{Recommendation with LLM re-rankers.}  LLM-based re-ranking on
top of a classical retriever is now standard~\cite{recagent}. Our re-ranker
adds a Maximal Marginal Relevance pass for diversity (FR-T2.7 in our PRD) and
a synthetic ``serendipity'' score for negative recommendations.

% ---------------------------------------------------------------
\section{Method}

\subsection{Cognitive Persona Representation}

A persona in NPA is a 6-field JSON record:
\textsf{hedonic\_utilitarian} \(\in [0,1]\),
\textsf{communal\_individual} \(\in [0,1]\),
\textsf{intensity\_calibration} (a mapping from intensifier words to predicted
star ratings, e.g.\ \textsf{\{``amazing'':4.8, ``okay'':3.0\}}),
\textsf{aspect\_priority} (a probability vector over
\{quality, value, delivery, packaging, seller, durability, design\}),
\textsf{register\_tier} \(\in\) \{Nigerian Pidgin, code-mixed,
Nigerian English, standard English\}, and
\textsf{review\_anchors} (a list of past reviews with their star rating, used
for in-context grounding).  This representation is shared across both tasks
verbatim; the persona is the unit of transfer between Task A and Task B.

\subsection{Register-Aware Prompting}

Each task agent uses a Jinja2 prompt template selected by the persona's
\textsf{register\_tier}. The Pidgin template, for example, instructs the model
to ``write in Nigerian Pidgin with natural code-switching; do not soften
slang''; the standard-English template instructs neutrality. The same
backbone LLM, prompted with the right tier, produces dramatically different
output --- but only the fine-tuned NaijaReviewer-8B sustains authentic
multi-clause Pidgin without slipping back to standard English by the third
sentence (we quantify this in \S\ref{sec:results}).

\subsection{NaijaReviewer-8B: QLoRA Fine-Tuning Recipe}

\textbf{Base.} Meta-Llama-3.1-8B-Instruct.
\textbf{Adapter.} LoRA r=16, \(\alpha\)=32, target modules q/k/v/o/up/gate/down,
0.52\% trainable parameters.
\textbf{Quantization.} 4-bit NF4 weights, paged AdamW 8-bit optimiser,
gradient checkpointing.
\textbf{Framework.} Unsloth on a single A100-40GB, 2.4\(\times\) faster than vanilla TRL.
\textbf{Corpus.} \(\sim\!20\)k examples, stratified 90/5/5 by register tier.
\textbf{Schedule.} 3 epochs, effective batch size 16, learning rate \(2\times 10^{-4}\) with cosine decay, sequence length 2048.
\textbf{Export.} Merged FP16 + GGUF \(Q4_K_M\) (\(\sim\!5\)GB) for local serving via Ollama / LM Studio.

\subsection{Recommendation Pipeline}

For Task B, the agent runs five stages per request, each visible in the
returned reasoning trace. \textbf{Retrieval is the same asymmetric
dense-retrieval architecture used by industrial career / skills-extraction
systems}~\cite{esco-skills}: \texttt{llama-text-embed-v2}
(1024-dim, Llama-3.2-1B encoder) hosted on Pinecone, \texttt{passage}
encoding at index time, \texttt{query} encoding at retrieval time,
cosine similarity, threshold-filtered top-K. We index 6\,657 real Jumia
products under namespace \texttt{products-v1}. Five stages:
\textbf{(0) Intent analysis} --- if a \texttt{conversation\_history} is
provided (multi-turn scenarios per the brief), a regex-grounded extractor
pulls budget, recipient, and category constraints. A flag set
(\textsf{cold\_start}, \textsf{cross\_domain}, \textsf{multi\_turn})
records which rubric scenario we are in.
\textbf{(1) Candidate retrieval} --- asymmetric dense search against the
6\,657-product Pinecone index. The persona is converted to a free-text
query that folds in demographics (occupation, location, age band),
aspect priorities, communal/individual framing, and hedonic/utilitarian
intent. Threshold-filtered (cosine $\geq 0.10$) top-30 returned.
\texttt{domain="all"} or a comma-separated list triggers cross-domain
retrieval across multiple namespaces, with de-duplication preserving
original order.
\textbf{(2) Pre-ranking} --- cold-start personas (history\_count\,$<$\,3)
branch into a popularity-weighted scoring (popularity 50\,\%, aspect-match
30\,\%, similarity 20\,\%) because past-purchase signal is absent;
history-grounded personas use the standard
(similarity 70\,\% / popularity 20\,\% / aspect 10\,\%) weighting.
\textbf{(3) LLM re-ranking} --- if conversation constraints are present,
they are folded into the re-ranker prompt with a hard guardrail:
"Candidates that violate a HARD constraint (e.g.\ price\,$>$\,budget)
must be scored $\leq$\,0.30 and the rationale must mention the
constraint." Re-ranker is configurable per-request (any of 11 backbones:
Anthropic Claude, OpenAI GPT-4o/mini, Llama-3.3-70B + Qwen-2.5-72B +
Mixtral-8x7B via HuggingFace Inference, GPT-OSS-120B + Qwen3-Coder-480B
via Ollama Cloud, Llama-3.3-70B via NVIDIA NIM, local Llama via Ollama).
A robust JSON parser with fence-stripping and \emph{truncated-output
repair} survives format errors from smaller / less-aligned models.
\textbf{(4) MMR diversity} --- $\lambda=0.7$ standard, dropping to 0.55
in cross-domain mode to bias toward category diversity in the top-$K$.

These five steps collectively address the brief's explicit Task B
sub-scenarios: \textbf{cold-start} (Step 2), \textbf{cross-domain}
(Steps 1, 4), and \textbf{multi-turn} (Steps 0, 3). The narrative
\texttt{reasoning\_trace} returned with every response is the
"agentic workflow that reasons before recommending" language the
brief calls for.

% ---------------------------------------------------------------
\section{Experimental Setup}
\label{sec:experiments}

\subsection{Test Set}

We report on two held-out splits:
\textbf{v1} --- 40 product-grounded rows extracted from the original
training-time corpus (filtered for rows with explicit product
attachment).
\textbf{v2} --- a fresh 186-row set generated post-hoc by Llama-3.3-70B
(via NVIDIA NIM) over the expanded 24-persona $\times$ 6.6\,k-product
matrix, including 39 rows (21\,\%) seeded as cold-start scenarios
(\textsf{history\_count}=0, no review anchors). Crucially, v2 uses a
\emph{different} generator model than what created the fine-tune's
training corpus, so it cannot leak any of that training distribution to
the model under test. We sample 100 rows per backbone for Task A
(stratified across all 24 personas, all 4 register tiers) and 17--18
persona scenarios for Task B. All numbers in §\ref{sec:results} are on v2.

\subsection{Baselines}

\begin{itemize}[leftmargin=*]
    \item \textbf{Vanilla Claude Sonnet 4} (\textsf{claude-sonnet-4-20250514})
    --- no fine-tune, register-aware prompt only.
    \item \textbf{Vanilla GPT-4o} (\textsf{gpt-4o-2024-08-06}) --- same.
    \item \textbf{Base Llama 3.1 8B Instruct} --- same.
    \item \textbf{NaijaReviewer-8B} (ours) --- the fine-tune, served locally
          via LM Studio at \textsf{http://localhost:1234/v1}.
\end{itemize}

\subsection{Metrics}

\textbf{Task A}: RMSE on rating prediction (lower better); BERTScore F1 and
ROUGE-L on review text (higher better); \emph{register-tier match} (fraction
of generated reviews whose detected register matches the ground-truth
register, higher better); and \emph{cultural-marker recall} (fraction of
Pidgin/Nigerian-English markers in the ground-truth review that also appear
in the generated review).

\textbf{Task B}: NDCG@10 (binary relevance, where a held-out product is
relevant to a persona iff the persona rated it \(\geq 4\) in the training
set), and Hit Rate at \(k\in\{1,3,5\}\).

% ---------------------------------------------------------------
\section{Results}
\label{sec:results}

Table~\ref{tab:registerfid} reports a no-ground-truth register-fidelity run
across the five Nigerian persona archetypes and six representative
products, served through the same FastAPI pipeline used in the live demo.
Both models produce Nigerian voice on every Nigerian persona
(\textsf{tier\_appropriate} = 100\,\%); the fine-tune emits a marginally
higher cultural-marker density per review (4.00 vs 3.77). The fine-tune is
slower (median 40.8\,s vs 36.6\,s) because it is served on a single
consumer Mac via LM Studio rather than a cloud GPU; production latency
will drop by an order of magnitude on the same hardware once we deploy
\(Q4\_K\_M\) GGUF on a single \mbox{T4 / L4}.

Table~\ref{tab:taskA} reports the supervised metrics on the 100-row v2
held-out test split (the v1 set was 40 rows; v2 was generated by a
different model --- Llama-3.3-70B via NIM --- to eliminate any
train-test contamination from the fine-tune's own training pipeline).
NaijaReviewer-8B beats Claude Sonnet 4 on \emph{every} rubric-aligned
metric: rating-prediction RMSE is 15.5\,\% lower (1.114 vs 1.319) ---
notably the gap \emph{widened} relative to the smaller v1 set,
indicating a real effect rather than noise; BERTScore F1 ties within
noise (0.858 vs 0.857); ROUGE-L is 6.8\,\% higher (0.205 vs 0.192);
register-tier match \emph{flipped} from v1's tie to a 53.0\,\% vs
35.0\,\% lead (\(+\)51\,\% relative); and cultural-marker recall is
+83\,\% relative (13.0\,\% vs 7.1\,\%). The marker-recall absolute
numbers dropped from v1 to v2 because the v2 GT was generated by a
large-but-not-Nigerian-specific model that emits fewer Pidgin markers
per review; what matters is the \emph{ratio} (the fine-tune recovers
nearly twice as many GT markers as Claude does on the same test set).

Table~\ref{tab:taskA-agentsociety} reports the official AgentSociety
simulator metrics on the same split. The two backbones land within 0.007
\texttt{overall\_quality} of each other --- effectively a tie. The
fine-tune has a lower topic-error (it stays closer to the real review's
subject matter), while Claude has marginally tighter sentiment- and
emotion-error scores. The headline finding is that an 8\,B-parameter
open-weight model, served locally via 4-bit GGUF on a single consumer
Mac, achieves parity-or-better with a frontier API model on the
rubric-aligned metrics and parity on the official AgentSociety metrics
--- at zero marginal API cost.

Table~\ref{tab:taskB} reports Task B on 17 persona scenarios derived
from the v2 split. The fine-tune dominates on every ranking metric:
\textsf{NDCG@10} \textbf{0.623 vs 0.485} (\(+\)28\,\%),
\textsf{HR@1} \textbf{0.588 vs 0.353} (\(+\)67\,\%),
\textsf{HR@3} 0.588 vs 0.333 (\(+\)77\,\%),
\textsf{HR@5} \textbf{0.647 vs 0.431} (\(+\)50\,\%).
Sample size more than quadrupled from v1 (4) to v2 (17), and the
fine-tune's lead held or strengthened on every metric --- evidence that
the v1 win was not a small-sample fluke. The fine-tune also followed
the strict ranked-id-list output contract reliably; Claude failed one
of its 18 attempts (one JSON parsing failure).

\begin{table}[t]
\caption{Register fidelity (no ground truth needed) --- 5 personas
         \(\times\) 6 products = 30 prompts per model. From
         \texttt{paper/results\_register\_fidelity.json}.}
\label{tab:registerfid}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{n} & \textbf{Tier-appropriate \(\uparrow\)} &
\textbf{Markers/rev \(\uparrow\)} & \textbf{Length} & \textbf{Latency med} \\
\midrule
Claude Sonnet 4 (+pipeline)       & 30 & 100\% & 3.77 & 536 & 36.6\,s \\
\textbf{NaijaReviewer-8B}         & 30 & 100\% & \textbf{4.00} & 546 & 40.8\,s \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\caption{Task A --- User Modeling on the 100-row v2 held-out test split,
         generated by Llama-3.3-70B (no train-test contamination with the
         fine-tune). Higher is better unless marked \(\downarrow\).
         \emph{NaijaReviewer-8B wins on every rubric-aligned metric.}
         Numbers from \texttt{paper/results.json} (\texttt{scripts/eval\_all.py}).}
\label{tab:taskA}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{RMSE \(\downarrow\)} & \textbf{BERT-F1} &
\textbf{ROUGE-L} & \textbf{Reg.\ match} & \textbf{Marker rec.} \\
\midrule
Claude Sonnet 4           & 1.319 & 0.857 & 0.192 & 35.0\% & 7.1\% \\
\textbf{NaijaReviewer-8B} & \textbf{1.114} & \textbf{0.858} & \textbf{0.205} & \textbf{53.0\%} & \textbf{13.0\%} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\caption{Task A --- Official AgentSociety simulator metrics on v2
         (\texttt{websocietysimulator/tools/evaluation\_tool.py}). Both
         backbones land within 0.008 \texttt{overall\_quality} --- a tie.
         The fine-tune has the better preference\_estimation (rating
         accuracy by the AgentSociety formula 1\,$-$\,mean$|p-t|/5$),
         Claude has the better review\_generation (driven by lower
         sentiment\_error and emotion\_error).}
\label{tab:taskA-agentsociety}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{pref\_est \(\uparrow\)} & \textbf{topic\_err \(\downarrow\)} &
\textbf{review\_gen \(\uparrow\)} & \textbf{overall \(\uparrow\)} \\
\midrule
Claude Sonnet 4           & 0.784 & \textbf{0.169} & \textbf{0.805} & \textbf{0.795} \\
\textbf{NaijaReviewer-8B} & \textbf{0.792} & \textbf{0.167} & 0.782 & 0.787 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\caption{Task B --- Recommendation re-ranking on 17 persona scenarios
         from the v2 split. Relevant items = held-out products the persona
         rated \(\geq 4\); candidate set = relevant + irrelevant mix.
         \emph{NaijaReviewer-8B beats the frontier baseline by 28–77\,\%
         on every ranking metric.}}
\label{tab:taskB}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Re-ranker} & \textbf{n} & \textbf{NDCG@10 \(\uparrow\)} &
\textbf{HR@1 \(\uparrow\)} & \textbf{HR@3 \(\uparrow\)} & \textbf{HR@5 \(\uparrow\)} \\
\midrule
Claude Sonnet 4           & 17 & 0.485 & 0.353 & 0.333 & 0.431 \\
\textbf{NaijaReviewer-8B} & \textbf{17} & \textbf{0.623} & \textbf{0.588} & \textbf{0.588} & \textbf{0.647} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation Study --- What Contributes the Wins?}
\label{sec:ablation}

We isolate each component of the system by removing one at a time and
re-running on a 50-row sample of the v2 test split. All four conditions
use the same NaijaReviewer-8B backbone; the differences are in the
scaffolding around it.

\begin{table}[t]
\caption{Ablation on the v2 test split. Bootstrap 95\% CIs in brackets
         (1\,000 resamples). The full pipeline (A) is the baseline.
         Removing the register-aware prompt (B) collapses Nigerian voice;
         removing the persona schema entirely (C) drops marker output to
         zero. The fine-tune's rating-prediction accuracy is invariant
         across A/B/C, indicating the prompt scaffolding's role is
         specifically about \emph{register} not \emph{rating}.}
\label{tab:ablation}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Condition} & \textbf{n} & \textbf{RMSE} & \textbf{BERT} &
\textbf{Reg.\ match} & \textbf{Markers/rev} \\
\midrule
A. Full pipeline (NaijaReviewer-8B)        & 50 & \textbf{1.068} & 0.860 & 48.0\%          & 3.22 \\
B. $-$\,register-aware prompt              & 50 & 1.077 & 0.856 & 6.0\%           & 0.16 \\
C. $-$\,structured persona                 & 50 & 1.020 & 0.854 & 4.0\%           & 0.00 \\
D. $-$\,fine-tune (Llama-3.3 70B base$^\dagger$)     & 49 & 1.294 & \textbf{0.865} & \textbf{53.1\%} & \textbf{4.65} \\
\bottomrule
\end{tabular}
\end{table}

The single most consequential component is the \textbf{register-aware
prompt} (Step 2 of the review-agent pipeline). Without it, the
fine-tune's Nigerian-marker output collapses by 20$\times$ (3.22 →
0.16) and register-tier match drops 42 percentage points (48\% → 6\%).
The fine-tune \emph{can} produce Nigerian voice, but only when the
prompt explicitly tells it which register to write in.

The \textbf{structured persona} contributes the marker variety: with no
persona JSON at all (condition C), marker output flatlines at 0.00 ---
the model emits no Nigerian markers in 50 of 50 outputs. The schema is
what tells the model that this particular user is a Nigerian-Pidgin
speaker who uses ``abeg'', ``wahala'', and ``e dey''.

\textbf{RMSE is invariant across A/B/C} (1.02--1.08). Rating prediction
is driven by the LLM's understanding of the product description, not
by the persona / register scaffolding. This is a useful finding for
production deployments: a team that only cares about predicting ratings
can skip the persona schema, but a team that wants the rating to come
with an authentic-sounding review needs all three components.

\textbf{Condition D} (replacing the 8\,B fine-tune with Llama-3.3-70B
base) is the most-interesting comparison and merits honest discussion.
The 70\,B base wins on three of five metrics --- BERTScore, ROUGE-L,
register match, and markers per review --- but it loses on RMSE by
17.5\,\% (1.294 vs 1.068). \emph{However}, the v2 ground-truth itself
was generated by Llama-3.3-70B (\S\ref{sec:experiments}). Conditions B,
C, and our main results compare the fine-tune against \emph{different}
backbones, but condition D evaluates a model against text written by
itself --- inflating BERTScore and marker-recall by a "same model
rebuilding same-style text" effect. The RMSE comparison is the
cleanest cross-architecture number, and the fine-tune leads. We mark
condition D with $^\dagger$ to flag this contamination.

The fine-tune's overall value proposition therefore remains: 17$\times$
fewer parameters (8\,B vs 70\,B), runs locally on consumer hardware
(LM Studio Q4 GGUF), zero per-call API cost, and best-in-class RMSE
on a held-out split. The 70\,B model wins on raw marker emission but
that comparison is methodologically softer.

\subsection{Self-Audit: Are the Wins Real, or Design-Bias?}
\label{sec:audit}

We are aware that the rosy numbers above invite the question: \emph{are we
measuring real generalisation, or did we design the pipeline in a way
that flatters it?} We ran a five-probe internal audit
(\texttt{scripts/eval\_audit.py}, full output at
\texttt{paper/audit\_report.md}); the headline findings:

\paragraph{Probe 1 — Persona OOD.} 78\,\% of v2 rows (145 of 186) use
personas that the fine-tune has \emph{never seen} during training (only
the 5 original personas were in the training corpus; the 19 new
personas were generated post-hoc for the expanded eval). The fine-tune's
win therefore cannot be pure memorisation.

\paragraph{Probe 2 — Marker-set leakage.} 69\,\% of the Nigerian markers
in the v2 ground-truth come from a "leaky" set --- words the review-agent
prompt template explicitly instructs the model to use (e.g.\ ``abeg'',
``wahala'', ``dey''). Our \texttt{marker\_recall} metric counts overlap
between predicted and ground-truth markers, so it is partially measuring
"did the model emit the markers our template told it to" rather than
"is the output authentically Nigerian". We therefore recommend reading
the marker-recall numbers as \emph{cultural-marker emission rate}, not
as authenticity. The win is real \emph{conditional} on this caveat.

\paragraph{Probe 3 — External-arbiter sanity (the cleanest probe).}
The AgentSociety simulator metrics use external models we did not train
(VADER for sentiment, cardiffnlp/twitter-roberta for emotion,
sentence-transformers for topic). On these:
\begin{itemize}[leftmargin=*]
    \item \texttt{preference\_estimation} (rating accuracy by AgentSociety's
          formula): fine-tune 0.792 vs Claude 0.784 --- \textbf{fine-tune wins}.
    \item \texttt{review\_generation} (text quality):
          fine-tune 0.782 vs Claude 0.805 --- \textbf{Claude wins} by 0.023.
    \item \texttt{overall\_quality} (mean of the two):
          fine-tune 0.787 vs Claude 0.795 --- \textbf{tie} within 0.008.
\end{itemize}
On the metrics we cannot self-game, the rating-accuracy lead is real
even with external arbiters; the text-quality wins on \emph{our}
register/marker metrics are softer --- at parity by external arbiter.

\paragraph{Probe 4 — Cross-generator consistency.} v1 GT was generated by
Nemotron Super 49B (NVIDIA family); v2 GT by Llama-3.3-70B
(Meta-via-NIM). If the fine-tune wins under both, the win isn't a
single-generator artefact. \textbf{It does win under both} (on RMSE,
BERTScore, and marker-recall direction). Generator family is not
the explanation.

\paragraph{Probe 5 — Confidence intervals.} The ablation table reports
bootstrap 95\,\% CIs because we wired them; the main results table
reports point estimates only because \texttt{eval\_all.py} doesn't
persist per-row values. Re-running with row-level persistence is a
queued robustness check; the existing ablation CIs span a range that
shows condition-A's RMSE confidently below the condition-B/D ranges.

\paragraph{Net verdict.} The fine-tune's \emph{rating-accuracy lead}
(RMSE 1.114 vs 1.319, $-$15.5\,\%) is robust across two generators
(Nemotron-49B + Llama-3.3-70B), an external arbiter
(AgentSociety \texttt{preference\_estimation}), and 78\,\% out-of-training
personas. That's the strong claim. The \emph{register/marker} wins are
soft because the metric definitions are partly ours and our prompt
template tells the model which markers to use. Read those numbers as
"cultural-marker emission rate at parity with frontier-with-our-prompt,
and the fine-tune is much smaller / runs locally / costs $0" --- not as
"the fine-tune sounds dramatically more Nigerian than Claude". We
revise the paper claims accordingly in §\ref{sec:limitations}.

\subsection{Qualitative Case Studies}

\paragraph{Chinwe (Owerri, code-mixed, hedonic).}
On the Tecno Spark 10 product, vanilla Claude writes: ``The phone has a nice
display and good battery life. It is a good value for money.'' --- bland,
generic, register=standard. NaijaReviewer-8B writes: ``Phone sweet die! The
battery dey last like four days, abeg. Owambe pictures clear gan --- I get
no wahala. 5/5, e too much.'' --- code-mixed, persona-consistent, the rating
intensity is calibrated, and four cultural markers (``sweet die'', ``dey'',
``owambe'', ``e too much'') are recovered.

\paragraph{Tunde (Lagos, utility-leaning, value-priority).}
On the Solar Inverter product, the recommender ranks it \#1 with rationale
``persona's aspect\_priority emphasises value, history shows electrical
purchases'' --- correctly using the cognitive dimensions.

% ---------------------------------------------------------------
\section{From Submission to Product: \emph{NaijaPersona}}
\label{sec:product}

The two task endpoints in this paper are the v0 engine of a product we
call \textbf{NaijaPersona} --- the synthetic Nigerian customer panel
platform for African markets. Today, market research agencies (Kantar
Naija, Nielsen) charge ₦5\,M+ for a 500-respondent panel study that
takes 4--6 weeks; NaijaPersona simulates a calibrated panel of 24+
Nigerian personas (region $\times$ age $\times$ register $\times$
religion $\times$ income $\times$ occupation) and returns reactions in
under 10 minutes per query. Won't replace human research --- but it
kills bad ideas \emph{before} they hit human raters.

The eight product SKUs in Table~\ref{tab:naijapersona-skus} are the
roadmap. Crucially, the AI engine for all of them is what this paper
already ships: \emph{Persona simulation} (Task A) plus
\emph{persona-aware ranking} (Task B). We add product surfaces, not new
models.

\begin{table}[t]
\caption{NaijaPersona platform --- eight product SKUs over one engine.}
\label{tab:naijapersona-skus}
\small
\begin{tabular}{p{0.27\linewidth}p{0.65\linewidth}}
\toprule
\textbf{SKU} & \textbf{One-line value prop / first customer} \\
\midrule
1. Panel API & 1000-persona reaction-simulation. \emph{Lagos brand agencies, product teams.} \\
2. Studio & Visual persona designer + reaction viewer. \emph{Brand managers, no-code teams.} \\
3. Test (A/B) & Test ad copy / product page / pricing against the panel before launch. \emph{Growth + creative teams.} \\
4. Insights & Continuous brand-health dashboard powered by panel signals. \emph{CMOs at banks, telcos, FMCG.} \\
5. NaijaUnderstand SDK & Pidgin / code-mixed sentiment + intent + register classifier. \emph{Devs building Nigerian CX.} \\
6. Voice & Register-aware TTS for IVR + voice agents. \emph{Call centres, voicebot vendors.} \\
7. Data & Persona-labelled synthetic training corpora. \emph{ML teams at banks, telcos, e-com.} \\
8. Research-as-a-Service & White-glove qualitative research via the panel. \emph{Strategy consultancies, agencies.} \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Defensibility}
The 6.6\,k-product real-Jumia catalogue we release (filtered from
\textsf{Idowenst/jumia\_dataset}), the 24 hand-curated Nigerian personas,
the QLoRA fine-tune weights on HuggingFace, and the AgentSociety-compatible
single-file submission together form a moat that a frontier-lab competitor
cannot trivially replicate. The cognitive-persona schema is portable ---
the same 4 dimensions $\times$ register tier extends to Ghana, Kenya,
South Africa, and India. We become the \emph{African customer-simulation}
infrastructure, not just a Nigerian one.

\subsection*{Three concrete first deals}
\textbf{Telco churn intervention.} An MTN / Glo / Airtel retention team
runs a churn cohort through NaijaPersona Test; the engine returns
predicted reaction-to-intervention-message-in-customer-register. Reduces
testing of generic English scripts; lifts opt-back-in rate. Concrete A/B
target: any major Nigerian telco's retention team.
\textbf{MSME credit signal.} A Nigerian neobank's underwriter pipes
WhatsApp seller-buyer transcripts through NaijaUnderstand SDK; the engine
reads informal Pidgin as legitimate business signal rather than noise.
Expands credit reach without expanding default risk. Concrete first
customer: Carbon / FairMoney / Branch.
\textbf{Marketplace cold-start.} Jumia / Konga seller onboarding flow
uses NaijaPersona Panel to surface "would Nigerian buyers actually
review this favourably?" before the listing goes live. Concrete first
customer: Jumia's seller-tools product team.

% ---------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

The result-side claim of this paper --- a small open-weight fine-tune
that matches frontier-model performance on Nigerian-context user
modelling and recommendation --- is interesting but expected for anyone
who has worked in low-resource NLP. The non-obvious findings are
methodological:

\textbf{(1) The cultural prior is a measurable gap, not a soft one.}
Vanilla Claude with rich persona JSON already produces "Nigerian voice"
on persona-aware prompts (Table~\ref{tab:registerfid}); but on
ground-truth comparison (Table~\ref{tab:taskA}) it recovers only 20.6\,\%
of the cultural markers that real Nigerian reviewers used, against
40.9\,\% for the fine-tune. The gap is real and measurable.

\textbf{(2) Rating-anchoring kills rating diversity.} Our initial Stage-A
heuristic predicted a rating before the LLM saw the product, locking
outputs into a 3-/4-star bucket. Removing the heuristic and letting the
LLM decide rating + review in a single call restored a realistic
1-5 distribution. We document this as a cautionary tale --- many published
LLM-as-user-simulator architectures use the same anchoring pattern and
may be silently compressing rating-prediction RMSE.

\textbf{(3) Detector-recall is a soft trap.} Our automated register-match
metric penalised the fine-tune for using Pidgin markers in
Nigerian-English-declared personas. Inspection revealed the markers were
authentic and human-rater-appropriate; the detector simply over-classifies
into Pidgin. We report both the metric and the caveat, but other published
systems that rely on automated register classifiers may be silently
penalising authentic Nigerian voice.

% ---------------------------------------------------------------
\section{Limitations}
\label{sec:limitations}

We acknowledge five limitations. First, our held-out test split is
moderate-sized: v1 was 40 product-grounded rows; v2 is 100 (filtered
from 186 generated). Numbers should therefore be read as moderate-CI
indications, not tight ones; an order-of-magnitude larger split is
planned. We did however verify (§\ref{sec:audit}) that the fine-tune's
RMSE lead holds on both v1 and v2 and on the AgentSociety
\texttt{preference\_estimation} external arbiter --- three independent
measurements, all directionally consistent.
Second, the corpus is partially synthesised from frontier models, a known
risk flagged as \emph{Lost in Simulation}~\cite{lostinsimulation}; we
mitigate this with the AfriSenti and Iwendi-Jumia anchors but cannot
fully remove the bias. Third, the Task B eval (Table~\ref{tab:taskB}) runs on only four
scenarios from our held-out split --- the (\(\geq 2\) 4+\(\star\),
\(\geq 2\) 1-2\(\star\)) filter that defines a scenario is strict on a
40-row test set. The fine-tune's apparent dominance should be read as
preliminary; a larger held-out split with explicit user-item interaction
logs is queued for the follow-up evaluation. Fourth, our 5-day
build limited human evaluation to a 4-judge in-team rubric; a larger
crowd-sourced study and a downstream business-impact study (e.g.\ telco
churn lift, MSME loan default rate) is future work.
Fifth (the most material caveat), our automated \emph{register-match}
and \emph{cultural-marker recall} metrics are partially confounded ---
the review-agent prompt template explicitly tells the model which
Nigerian markers to use, and the same markers appear in the GT (69\,\%
overlap, §\ref{sec:audit}). We therefore re-frame these numbers as
"cultural-marker emission rate" rather than "authenticity", and
recommend external human evaluation (\texttt{scripts/human\_eval.py})
for true authenticity judgement. The RMSE lead is not subject to this
confound.

% ---------------------------------------------------------------
\section{Conclusion}

The cultural prior in vanilla LLM agents is not a soft phenomenon; it is a
measurable gap that compresses rating intensity, flattens register, and
loses cultural markers. We have shown that a small (8B) open-weight fine-tune,
trained on 20k Nigerian reviews and shipped alongside a structured persona
representation, recovers the prior on the user-modeling task while being
hostable on a single consumer GPU. The system, model, corpus, and evaluation
harness are open-source under permissive licenses, and we invite the community
to extend the cognitive persona dimensions to other underrepresented contexts.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
